---
title: "Assignment_01"
output: html_document
date: "2025-01-15"
---

```{r}
BirdMove <- read.csv("http://gannet.fish.washington.edu/seashell/snaps/BirdMove.csv")
```
```{r}
GPS22 <- read.csv("http://gannet.fish.washington.edu/seashell/snaps/GPS.csv")
```

Question 2
In this question we will create a line graph of the distances travelled by the birds through time.
Firstly, the date column in our data isn’t formatted as a date (by default R reads in strings of text as just
that, a string of text). However, we can re-format those strings as a date using the as.Date() function by
following the commands below
# The following code converts the date column to a date class
# The format argument tells R how to interpret the string of txt
BirdMove$date <- as.Date(BirdMove$date, format="%d/%m/%Y")


```{r}
BirdMove$date <- as.Date(BirdMove$date, format="%d/%m/%Y")
```



```{r}
# Initial plot for the first bird
plot(BirdMove$BirdID_2, 
     xlab = "Date", 
     ylab = "Daily travel distance (km)", 
     main = "Bird 2 travel distance (km) from March 2019-August 2019", 
     ylim = c(0, 250), 
     lwd = 2, 
     type = "l")

# Add a second line for another bird (e.g., BirdID_3)
lines(BirdMove$BirdID_3, 
      col = "red", # Choose a different color for distinction
      lwd = 2, 
      type = "l") # Same line type
```






Copy this code above into your R session and make sure that the dates look correct (they should appear
as YYY-MM-DD).
2a) Create a line graph of travel distances through time for bird ID_2, plotting daily travel distance
against date.
When creating the plot add the following plot attributes by specifying the necessary arguments in the
plot function (for more help with the plot function type ?plot in R):
● The default plot type in R is a scatterplot, we want a line graph instead. To do this, use the
type="l" option.
● Label the x axis, using the xlab argument, “Date”
● Label the y axis, using the ylab argument, “Daily travel distance (km)”
● Change the y-axis range using the ylim argument so that the axis ranges from 0 to 250
● Add a plot title using the main argument. Set the plot title to be a brief description of the
information plotted followed by your name in parentheses
● Change the line width using the lwd function to 2
Take a screenshot of your graph and paste it below (6 pts)
2b) Now overlap the distances travelled by Bird ID_3 onto this figure using the lines() function, changing
the color of the Bird ID_3 line to distinguish it from Bird ID_2, but keeping the same linewidth as before.
Once the 2nd line is added, add a legend to tell the viewer which line belongs to which bird using the
legend() function. (4 points)
2c) Comparing the distances travelled, provide a brief description of any major differences you see
between the two patterns through time (2 pts)







```{r}
# Create a histogram with labeled x and y axes
hist(GPS$HDOP, 
     main = "Histogram of HDOP Values",  # Add a title
     xlab = "HDOP",                     # Label for the x-axis
     ylab = "Frequency")                # Label for the y-axis

```

